The error when inferring phylogenies with incipient species by a Birth-Death model

### 2018-04-09 article.tex
Add 'How to construct a Nature summary paragraph'
Use @rsetienne's suggested introduction

@@

% Use 'It was found that A is B (Name 1234)' style
\setcitestyle{authoryear,open={},close={}}

@@

  The tools for reconstructing phylogenetic relationships between taxonomic units (e.g. species) have become very advanced in the last two decades. Among the most popular tools are Bayesian approaches, such as BEAST, MrBayes and RevBayes, that use efficient tree sampling routines to provide a posterior probability distribution of the phylogenetic tree. A feature of these approaches is the possibility to incorporate known or hypothesized structure of the phylogenetic tree through the tree prior. It has ebe shown that the effect of the prior on the posterior distribution of trees can be substantial. Currently implemented tree priors assume that speciation is instantaneous. However, we know that speciation can be a gradual process and this idea has been termed protracted speciation. Here we explore the effects of ignoring the protractedness of the speciation process with an extensive simulation study. We simulate protracted phylogenies using the protracted birth-death process and then simulate sequence alignments according to a substitution model and clock model. We then use BEAST to infer the phylogeny from these alignments using the standard birth-death prior that assumes instantaneous speciation. We compare the inferred tree to the simulated tree, and find that ....

  Furthermore, we identify an important issue related to protracted speciation. Because the tree produced by the protracted birth-death process is not necessarily monophyletic, we cannot speak of "the" species tree. Instead we have to sample among the incipient species to represent species. We show that different ways of sampling the representative species results in ....

  We conclude that ...

@@

{\bf Keywords:} computational biology, evolution, phylogenetics, BEAST2, R

@@

Phylogenetics enables us to infer which species lived when and
estimate how long it took for a species to give rise to a new
lineage. Phylogenies show us taxa that have completed speciation, yet
those taxa only: species that are in the process of speciation are 
overlooked.

There are multiple reasons why a speciation event, during
or after its course, goes unnoticed. It may be that the build-up
of reproductive isolation is still in an intermediate stage. But it
may also be that speciation has completed, but we, humans, have not
yet observed it. One example are the four African giraffe species, 
that have been thought to have been one single species up until 2016,
where the MRCA of the two most recent species goes back 1.6 million years.

It took some time until the Birth-Death model, a phylogenetic model
assuming constant rates of speciation and extinction, got extended
by a version that incorporates the fact that speciation takes time.
This extension is called the Protracted Birth-Death model. In this
model, a branching event does not give rise to a new species, but to
a new species-to-be, called an incipient species. Such an incipient
species may go extinct, or finish its speciation to become a good species.

Although we know speciation takes time, most of our theoretical 
phylogenetic modeling ignores this fact. To the best of our knowledge,
there is no tool incorporating this fact. One reason may be to ease
computation, enabling researchers to work on bigger phylogenies (with
more power statistical power) instead. Additionally, speciation being
protracted is only playing a role in the present. As seen from the present,
all older species have finished speciation, as we -by definition- recognize
these species as such.

There are problems with ignoring protracted speciation. A very pragmatic
example is the conservation of the now four smaller groups of giraffes,
instead of one big group. But from a more theoretical perspective, the
observed decline of speciation rate in time (excluding the contemporary human-driven
mass extinction event) may be explained by these new species not being 
recognized yet. Another problem is that the MRCA of young species is estimated 
incorrectly, as in the BD model a species can be ultra-young (as speciation
is assumed to be instantaneous), where the PBD model will place these MRCAs
further back in time (as two events need to have finished completion).

@@

This exploratory research's goal is to produce a 
usefull and balanced data set, that quantifies the inference error 
we make from choosing an overly simplistic prior. In this context, 
we measure the error made on phylogenies in which speciation takes 
time, when using a Birth-Death prior.

@@

From an alignment, we run a Bayesian analysis and create a posterior, 
using the phylogetic tool BEAST2. We assume a
JC69 nucleotide substitution (as we do use that), a strict clock model (as
we have that) and a Birth-Death tree prior (as this simplification is the goal
of this research). Additionally, we assume a MRCA prior with a normal distribution
with a mean of the crown age, and a standard deviation of 0.001 time units. 0.001
time units equates to 1000 years, which is the same resolution as which the
alignment is created. After discarding a burn-in of 10\%, the MCMC algorithm must 
result in an ESS of the posterior of at least 200 and the MCMC chain length is 
chosen as such that this is the case for all simulations. For simulations
with an ESS below 200, their MCMC chain length is elongated. An
MCMC chain is elongated by the proportion that is expected to increase it to 220. 
For example, would an ESS be 165, increasing its MCMC chain length by 33\% 
would bring it to (165\%*1.33=) 220. We use 220 as a safeguard to prevent 
an asymptotic approach to 200.

@@

Each posterior's phylogeny is compared to the 'true' species tree.
To quantify the difference between two phylogenies, the nLTT statistic
is used. The nLTT statistic equates to the area between the normalized
lineages-through-time-plots of two phylogenies, which will be a value between 
zero (for identical phylogenies) and one. As the nLTT statistic increases
when phylogenies get increasingly different, we use the nLTT statistic
as a measure of inference error. Comparing the one 'true' species tree
with each of the posterior's species trees, a distribution nLTT statistics
is created.

@@

To visualize the data set, we will produce two figures.
The first figure will display the distribution 
of median nLTT statistics per combination of speciation initiation, speciation
completion and extinction rate. We will use the median nLTT statistic over 
summing nLTT statistic distributions, as we think this will give the cleanest
visualization [NOTE: I am unconvinced, wouldn't a violin plot would
be better?]. We will use the median nLTT statistic, instead of the mean, as
we expect the distribution of nLTT statistics not to be symmetrical. The distribution
of median nLTT statistics is displayed per biological parameter set in a box-plot,
to shown of that distribution as needed, without showing an overly amount of detail
regarding that distribution. 
The second figure will show the distribution of median nLTT statistics per expected
mean duration of speciation. The expected mean duration of parameter can be calculated
with equation 19 of [Etienne \& Rosidell 2012] and is a combination of the three
biological parameters combined to their relevant biological effect. Through these
points, we'll fit a straight line (we have no reason to expect an effect of
higher order) and a locally weighted scatterplot smoothing (LOESS) 
curve, which is a non-parametric fit to expose potential higher order effects.
Additionally, we'll draw a horizontal line to the average median nLTT value 
[NOTE: here the median of all distributions would feel better] of
those simulations with instantanous speciation, which serves as a baseline of
the inference error made.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}
  \centering 
  \begin{tabular}{l l l}
    \hline
    Parameter             & Description & Values \\
    \hline
    \hline
    $b = b_g = b_i$       & Speciation initiation rate & 0.1, 0.5, 1.0 \\
    $\lambda$             & Speciation completion rate & 0.1, 0.3, 1.0, $10^6$ \\
    $\mu = \mu_g = \mu_i$ & Extinction rate & 0.0, 0.1, 0.2, 0.4 \\
    $t_c$                 & Crown age & 15 \\
    $\sigma_c$            & Standard deviation around crown age & 0.001 \\
    M                     & Sampling method & Youngest, oldest \\
    $r$                   & Mutation rate & 1000 \\
    $l_a$                 & DNA alignment length & $15000$ \\
    L                     & MCMC chain length & 1M \\
    $R_i$                 & RNG seed incipient tree & 1 to 40000 \\
    $R_a$                 & RNG seed alignment simulation & 1 to 40000 \\
    $R_b$                 & RNG seed BEAST2 & 1 to 40000 \\
    \hline
  \end{tabular}
  \caption{
    Overview of the 12 simulation parameters.
  }
  \label{table:simulation_parameters}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}
  \centering 
  \begin{tabular}{l l}
    \hline
    $n$ & Description \\
    \hline
    \hline
    $12$   & simulation parameters \\
    $1000$ & nLTT statistic \\
    \hline
    \hline
    $12$   & simulation parameters \\
    $1$ & [NOTE: add yes/no?] expected speciation completion time (from Etienne \& Rosindell 2012) \\
    $1$ & [NOTE: add yes/no?] expected duration of speciation (from Etienne \& Rosindell 2012) \\
    $1$ & number of taxa in incipient species tree (also: number of subspecies) \\
    $1$ & [NOTE: add yes/no?] difference in branch lengths between the two sampling methods? \\
    $1$ & number of taxa in species tree (also: number of species) \\
    $1$ & mean nLTT statistic \\
    $1$ & median nLTT statistic \\
    $1$ & lower bound of 95\% HPD value of nLTT statistic \\
    $1$ & upper bound of 95\% HPD value of nLTT statistic \\
    $11$ & ESSes of all parameters estimated by BEAST2 (see specs below) \\
    $?10$ & MCMC operator acceptances\\
    \hline
  \end{tabular}
  \caption{
    Specification of the data sets. $n$ denotes the number
    of columns a certain item will occupy. Above the double horizontal line,
    the content of the raw data file is specified, resulting in a table of 
    1012 columns and 40K rows. Below the double horizontal line,
    the content of the summary data file specified, resulting in a table of
    42 columns and 40K rows.
  }
  \label{table:raw_data_set_specs}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

% Short communications of approximately 3000 words are also accepted. 
% These papers should contain no more than two figures, two tables, and thirty references. 
% A short abstract of fewer than 200 words is acceptable.

@@

% Use 'It was found that something is something (Name 1234)' style
\setcitestyle{authoryear,open={},close={}}

@@

  % From 'How to construct a Nature summary paragraph'

  % A short abstract of fewer than 200 words is acceptable.

  % One or two sentences providing a basic
  % introduction to the field,
  % comprehensible to a scientist in any discipline.
  The tools for reconstructing phylogenetic relationships between taxonomic 
  units (e.g. species) have become very advanced in the last two decades.

  % Two to three sentences of
  % more detailed background, comprehensible to
  % scientists in related disciplines.
  Among the most popular tools are Bayesian approaches, such as BEAST, MrBayes and RevBayes, 
  that use efficient tree sampling routines to create a posterior probability distribution 
  of the phylogenetic tree. 
  A feature of these approaches is the possibility to incorporate 
  known or hypothesized structure of the phylogenetic tree through the tree prior. 
  It has been shown that the effect of the prior on the posterior distribution 
  of trees can be substantial.

  % One sentence clearly stating the general
  % problem being addressed by this particular
  % study.
  Currently implemented tree priors assume that speciation is instantaneous,
  where we know that speciation can be a gradual process.

  % One sentence summarising the main
  % result (with the words “here we show”
  % their equivalent).
  Here we explore the effects of ignoring 
  the protractedness of the speciation process with an extensive simulation study.

  % Two or three sentences explaining what
  % the main result reveals in direct
  % comparison to what was thought to be the case
  % previously, or how the main result adds to
  % previous knowledge.

  % One or two sentences to put the results into a
  % more general context.

  % Two or three sentences to provide a
  % broader perspective, readily comprehensible
  % to a scientist in any discipline, may be included
  % in the first paragraph if the editor considers that the accessibility of the paper is significantly enhanced
  % by their inclusion. Under these circumstances, the length of the paragraph can be up to 300 words.
  % (The above example is 190 words without the final section, and 250 words with it).
  We conclude that ...

  % Does not seem to fit here: 
  %  
  % We simulate protracted phylogenies using the protracted birth-death process 
  % and then simulate sequence alignments according to a substitution model and clock model. 
  % We then use BEAST to infer the phylogeny from these alignments using the standard 
  % birth-death prior that assumes instantaneous speciation. We compare the 
  % inferred tree to the simulated tree, and find that ....
  % 
  % Furthermore, we identify an important issue related to protracted speciation. 
  % Because the tree produced by the protracted birth-death process 
  % is not necessarily monophyletic, we cannot speak of "the" species tree. 
  % Instead we have to sample among the incipient species to represent species. 
  % We show that different ways of sampling the representative species results in ....

@@

{\bf Keywords:} computational biology, evolution, phylogenetics, prior choice

@@

The computational tools a contemporary phylogeneticist has at his/her disposal
goes beyond the wildest imagination of those living two decades ago. 
Advances in computational power allowed the first cladograms to be inferred 
from DNA alignments in [some year, some ref]. The first time-trees were
published in [some year, some ref]. Thanks to the addition of effective sampling
methods, the first Bayesian tools appeared in [some year, some ref], allowing
for unprecedented flexibility in the setup of a phylogenetic model.

Currently, the most popular Bayesian phylogenetics tools are BEAST [ref],  
MrBayes [ref] and RevBayes [ref]. Each of these use efficient tree sampling 
routines to create a posterior probability distribution of the to-be-inferred 
phylogenetic tree. But next to the computational speed, they also allow
to incorporate known or hypothesized structure of the phylogenetic 
tree-to-be-inferred through the model priors.

The model prior most relevant here, is the tree prior. The tree prior
embodies the speciation model assumed. Each speciation model has a
different expected branch length distribution. It has been shown 
[NOTE: where?] that the effect of the prior on the posterior distribution 
of trees can be substantial.

All of the currently implemented tree priors assume that speciation is 
instantaneous, where we know that speciation can be a gradual process.
The protracted birth-death (PBD) model, an extension of 
the (constant-rate) birth-death (BD) model, incorporates this idea.
In the PBD model, a branching event does not give rise to a new species, but to
a new species-to-be, called an incipient species. Such an incipient
species may go extinct, or finish its speciation to become a good species.

Although we know speciation takes time, all of our theoretical 
phylogenetic modeling tools ignore this fact. One reason may be to ease
computation, enabling researchers to work on bigger phylogenies (with
more power statistical power) instead. Additionally, speciation being
protracted is only playing a role in the present. As seen from the present,
all older species have finished speciation, as we -by definition- recognize
these species as such.

There are problems with ignoring protracted speciation. A very pragmatic
example is the conservation of an underestimated amount of species. 
But from a more theoretical perspective, the
observed decline of speciation rate in time (excluding the contemporary human-driven
mass extinction event) may be explained by these new species not being 
recognized yet. Another problem is that the MRCA of young species is estimated 
incorrectly, as in the BD model a species can be ultra-young (as speciation
is assumed to be instantaneous), where the PBD model will place these MRCAs
further back in time (as two events need to have finished completion).

@@

This exploratory research's goal is to produce a 
useful and balanced data set, that quantifies the inference error 
we make from choosing the incorrect prior. In this context, 
we measure the error made on PBD phylogenies, when using a BD prior.

@@

From an alignment, we run a Bayesian analysis and create a posterior, 
using the phylogetic tool BEAST2. We assume a
JC69 nucleotide substitution (as we do use that), a strict clock model (as
we have that) and a Birth-Death tree prior (as this simplification is the goal
of this research). Additionally, we assume a MRCA prior with a normal distribution
with a mean of the crown age, and a standard deviation of 0.001 time units. 0.001
time units equates to 1000 years, which is the same resolution as which the
alignment is created. After discarding a burn-in of 10\%, the MCMC algorithm must 
result in an ESS of the posterior of at least 200 and the MCMC sampling interval is 
chosen as such that this is the case for all simulations. For simulations
with an ESS below 200, their MCMC sampling interval is doubled up until
the ESS exceeds 200.

@@

Each posterior's phylogeny is compared to the (sampled) species tree.
To quantify the difference between two phylogenies, the nLTT statistic
is used. The nLTT statistic equates to the area between the normalized
lineages-through-time-plots of two phylogenies, which will be a value between 
zero (for identical phylogenies) and one. As the nLTT statistic increases
when phylogenies get increasingly different, we use the nLTT statistic
as a measure of inference error. Comparing the one (sampled) species tree
with each of the posterior's species trees, a distribution nLTT statistics
is created.

@@

To visualize the data set, we will produce two figures.
The first figure will display the distribution 
of median nLTT statistics per combination of speciation initiation, speciation
completion and extinction rate. We will use the median nLTT statistic over 
summing nLTT statistic distributions, as we think this will give the cleanest
visualization [NOTE: I am unconvinced, wouldn't a violin plot would
be better?]. We will use the median nLTT statistic, instead of the mean, as
we expect the distribution of nLTT statistics not to be symmetrical. The distribution
of median nLTT statistics is displayed per biological parameter set in a box-plot,
to shown of that distribution as needed, without showing an overly amount of detail
regarding that distribution. 
The second figure will show the distribution of median nLTT statistics per expected
mean duration of speciation. The expected mean duration of parameter can be calculated
with equation 19 of [Etienne \& Rosidell 2012] and is a combination of the three
biological parameters combined to their relevant biological effect. Through these
points, we'll fit a straight line (we have no reason to expect an effect of
higher order) and a locally weighted scatterplot smoothing (LOESS) 
curve, which is a non-parametric fit to expose potential higher order effects.
Additionally, we'll draw a horizontal line to the average nLTT 
value (over all replicates) of those simulations with instantaneous speciation, 
which serves as a baseline of
the inference error made.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}
  \centering 
  \begin{tabular}{l l l}
    \hline
    Parameter             & Description & Values \\
    \hline
    \hline
    $b = b_g = b_i$       & Speciation initiation rate & 0.1, 0.5, 1.0 \\
    $\lambda$             & Speciation completion rate & 0.1, 0.3, 1.0, $10^6$ \\
    $\mu = \mu_g = \mu_i$ & Extinction rate & 0.0, 0.1, 0.2, 0.4 \\
    $t_c$                 & Crown age & 15 \\
    $\sigma_c$            & Standard deviation around crown age & 0.001 \\
    $M$                   & Sampling method & Youngest for odd $R_i$, oldest for even $R_i$ \\
    $r$                   & Mutation rate & 1K \\
    $l_a$                 & DNA alignment length & $15K$ \\
    $n_s$                 & Number of MCMC states & 1K \\
    $f_i$                 & MCMC sampling interval & 1K or more \\
    $R_i$                 & RNG seed incipient tree & 1 to 40K \\
    $R_a$                 & RNG seed alignment simulation & 1 to 40K \\
    $R_b$                 & RNG seed BEAST2 & 1 to 40K \\
    \hline
  \end{tabular}
  \caption{
    Overview of the 12 simulation parameters.
  }
  \label{table:simulation_parameters}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}
  \centering 
  \begin{tabular}{l l}
    \hline
    $n$ & Description \\
    \hline
    \hline
    $12$   & simulation parameters \\
    $1000$ & nLTT statistic \\
    \hline
    \hline
    $12$ & simulation parameters \\
    $1$ & expected duration of speciation (from Etienne \& Rosindell 2012) \\
    $1$ & number of taxa in incipient species tree (also: number of subspecies) \\
    $1$ & number of taxa in species tree (also: number of species) \\
    $1$ & mean nLTT statistic \\
    $1$ & median nLTT statistic \\
    $1$ & lower bound of 95\% HPD value of nLTT statistic \\
    $1$ & upper bound of 95\% HPD value of nLTT statistic \\
    $11$ & ESSes of all parameters estimated by BEAST2 (see specs below) \\
    \hline
  \end{tabular}
  \caption{
    Specification of the data sets. $n$ denotes the number
    of columns a certain item will occupy. Above the double horizontal line,
    the content of the raw data file is specified, resulting in a table of 
    1012 columns and 40K rows. Below the double horizontal line,
    the content of the summary data file specified, resulting in a table of
    42 columns and 40K rows.
  }
  \label{table:raw_data_set_specs}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
